--- 
title: "Set up R-scripts to run automatically"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Set up R-scripts to run automatically}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## Overview
Automatic running of R-scripts at scheduled times is useful for making routine reports that should be run regularly, for example outbreak reports and extracting data for statistics like the "Skrantesykestatikk". This vignette describes the different steps to set up automatic batch jobs at Windows.

1. [Set up Windows Task Scheduler to run a bat-script](#set-up-windows-task-scheduler)
2. [Write the bat-script that run the R-scripts](#bat-script)
3. [Write the main R-script](#main-r-script)
4. [Write the R-script to check the log](#r-script-to-check-the-log)

## Set up Windows Task Scheduler
You can use the Task Scheduler in Windows to run an R-script automatically. To set up the Task Scheduler, see [Task Scheduler help](https://www.windowscentral.com/how-create-automated-task-using-task-scheduler-windows-10). Contact epi to get access to a server to host the script.

## bat-script
Use a bat-file to call R-scripts. We call two R-scripts, the main script that sources all R-scripts to do the job,  and a R-script to check the log after the main script ws run and report the status after running the main script. 

A typical bat-file can look like this:
```
"C:\Program Files\R\R-4.0.3\bin\x64\R.exe" CMD BATCH "\\fullpath\batch_outbreak_123.R"
"C:\Program Files\R\R-4.0.3\bin\x64\R.exe" CMD BATCH "\\fullpath\batch_outbreak_123_check_log.R"
```

Be aware that if the path names to the R-scripts include spaces, the path name must be enclosed in quotes. 
National characters should be avoided in the path name. If they cannot be avoided and the bat-file cannot start the R-scripts, try the following. 
Identify the code oge at the Windows server running the bat-file. 
Start the bat-file with setting the codepage to the same as the code page for the server where the R-scripts are located. 
End the bat-file by setting the code page back to the original code page. 

## Main R-script
Give the main script a stable name that is self explanatory. The name of the main script is included in the bat-file command and you prefer to not have to change the bat-file command. An example is batch_outbreak_123_disease_2020.R

The main script sources different R-scripts that does the tasks. The batch script will usually consist of three parts, setting up the R-environment, sourcing R-scripts, and making the end of the script. 

Setting up the R-environment
Attach packages
```{r, include = TRUE} 
library(NVIbatch) 
use_pkg(pkg = c()) 
use_NVIverse(pkg = c()) 
```

Global variables
```{r, include = TRUE} 
today <-
``` 

Source R-scripts 
```{r, include = TRUE} 

``` 
marking the end of the script
```{r, include = TRUE} 

``` 

## R-script to check the log
The aim of this script is to perform a check if the main script did run, and if not, report the error and warning messages. The script generates an email that is sent to specified persons.

Give the log script a stable name that is self explanatory. The name of the log script is included in the bat-file command and you prefer to not have to change the bat-file command. I prefer to give a name that is built on the main script name, for example batch_outbreak_123_disease_2020_check_log.R.

The check log script will usually consist of five parts, setting up the R-environment, save the log in the archive, check the log for warnings and errors, report the result of the check per email and delete old log files in the archive.

Setting up the R-environment
Attach packages
```{r, include = TRUE} 
library(NVIbatch) 
use_pkg(pkg = c("sendmailR")) 
use_NVIverse(pkg = c("NVIdb")) 
```

Global variables
```{r, include = TRUE} 
# Global variables
# Paths and filenames
domene <- paste0(set_dir_NVI("FAG"), "xxxx")

batch <- "batch_outbreak_123_disease_2020" #= navnet på scriptet som det skal lage logg for

#Dagens dato i formatet yyyymmdd til bruk i filnavn
today <- format(Sys.Date(),"%Y%m%d")
```

Save log in archive
```{r, include = TRUE} 
# SAVE LOG-FILE PER DAY ----
file.copy(from = paste0(domene,"Rscripts/",batch,".Rout"), 
          to = paste0(domene, "batchlogg/", batch, " ", today, ".Rout"))

``` 

Check log for errors and warnings
```{r, include = TRUE} 
messages <- gather_messages(filename = paste0(domene, "Rscripts/", batch, ".Rout"))
``` 

Send email with the results of the check
```{r, include = TRUE} 
#needs full path if not in working directory
attachmentPathName <- paste0(domene, "Rscripts/", batch, ".Rout")
#same as attachmentPath if using working directory
attachmentName <- paste0(batch, ".Rout")
#key part for attachments, put the body and the mime_part in a list for msg
attachmentObject <- mime_part(x = attachmentPathName, name = attachmentName)

# Tar inn feilmeldinger dersom det er noen
if (length(messages)>0) {
  subject <- paste("Feil i", batch)
  body <- list(c("Feilmeldinger", messages), attachmentObject)
} else {
  subject <- paste("Status for", batch)
  body <- list(attachmentObject)
}

# SEND EMAIL ----
sendmail(from = "<epi@vetinst.no>",
         to = c("<Petter.Hopp@vetinst.no>"),
         subject = subject,
         msg = body, 
         control = list(smtpServer="webmail.vetinst.no"))
``` 

Delete archived files
```{r, include = TRUE} 

```

